---
title: "Exercise 3 --- Point Patterns in Space and Space-Time"
subtitle: "BSTA 374 · 25 points · Due Sunday, October 4, 2026, 11:59 pm"
author: "Anik Das · Lehigh University"
abstract: "Published page: <https://dasanikbd.github.io/BSTA-374/Exercise3/DAS_A_Exercise3.html>"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
  pdf:
    papersize: letter
    geometry:
      - margin=1in
    toc: false
    colorlinks: true
    code-overflow: wrap
prefer-html: true
execute:
  echo: true
---

## What this exercise is

The largest lab of the semester. You will decide whether a set of case locations is clustered, at what spatial
scale, where the intensity is highest, and whether the clustering is also
concentrated in *time* --- which is a different claim, and the one that matters
for an outbreak.

**Data.** Two files, posted with this exercise. Put them in a `data/` folder
beside your `.qmd`.

| File | What it holds |
|---|---|
| `dengue_cases_sim.csv` | 500 simulated dengue cases: `case_id`, `onset_date`, `report_date`, `neighborhood`, `age_group`, `lon`, `lat` |
| `study_area.geojson` | the study-area boundary polygon |

These are **simulated** data modelled on a dengue surveillance setting. Nobody's
health information is involved, and the generating process is known --- which
means your conclusions can be checked.

Each part ends with **starting code**: copy it, fill in the blanks marked
`______`, and adapt. It is scaffolding, not a solution.

::: {.callout-important}
## Submission
Two things. Submit `LASTNAME_FI_Exercise3.qmd` --- the source --- to Course Site,
with your name in the document header. And publish the rendered HTML to your
semester GitHub repository, then paste the live URL into the Course Site
submission --- Part H is that second step, and it is the one that carries the
point.

**Due Wednesday, October 1, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Restart R and render from a clean session before submitting. Several parts
involve simulation --- set a seed at the top and use it, or your numbers will
change between the render you checked and the render I grade.
:::

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(spatstat)
library(splancs)      # space-time K function
library(plotly)       # space-time cube

set.seed(800200879)      # My student ID 
```

---

## Part A --- From CSV to `ppp` (3 points)

Everything after this part runs on a `ppp` object, so it is worth knowing what
one is. `spatstat` does not work with `sf`. A **`ppp`** ("planar point pattern")
holds two things: the coordinates of the events, and a **window** --- the region
within which they could have been observed. Both are required, because every
statistic in this lab is a comparison against what would be expected inside that
window: the intensity is $n$ divided by its area, the expected nearest-neighbour
distance depends on that intensity, and the edge corrections depend on its shape.
Give `spatstat` the wrong window and every number below changes without any
warning.

1. Read the case file and the boundary. Report how many cases you have and what
   one row represents.
2. Convert the cases to an `sf` object in **EPSG:4326**, then project both the
   cases and the boundary to **EPSG:32618** (UTM 18N, metres). One sentence on
   why the analysis cannot be done in degrees.
3. Build a `ppp` whose window is the study-area polygon, then print it. Report
   the number of events and the intensity in cases per km².
4. Replace the polygon window with the bounding box of the cases
   (`Window(pp) <- owin(range(x), range(y))`) and recompute the intensity. Two or
   three sentences on how much it changed and why the choice of window is a
   modelling decision rather than a technicality. Then go back to the polygon.
5. Map the cases over the boundary.

### Starting code

```{r}
#| label: import-and-window
cases_raw <- read.csv("data/dengue_cases_sim.csv", stringsAsFactors = FALSE)
area <- st_read("data/study_area.geojson", quiet = TRUE) |>
  st_transform(32618)
cases <- cases_raw |>
  mutate(onset_date = as.Date(onset_date),
         report_date = as.Date(report_date)) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(32618)

stopifnot(nrow(cases) == 500L,
          !anyNA(cases$onset_date),
          all(st_covered_by(cases, st_union(area), sparse = FALSE)))
xy <- st_coordinates(cases)
win <- as.owin(st_geometry(area))
pp <- ppp(xy[, 1], xy[, 2], window = win)
pp

poly_km2 <- area.owin(win) / 1e6
poly_intensity <- intensity(pp) * 1e6
pp_box <- pp
Window(pp_box) <- owin(xrange = range(pp$x), yrange = range(pp$y))
box_intensity <- intensity(pp_box) * 1e6
window_summary <- data.frame(
  window = c("Study-area polygon", "Case bounding rectangle"),
  area_km2 = c(poly_km2, area.owin(Window(pp_box)) / 1e6),
  cases_per_km2 = c(poly_intensity, box_intensity)
)
knitr::kable(window_summary, digits = 2)                  
```
```{r}
#| label: case-map
#| fig-width: 7
#| fig-height: 6
ggplot() +
  geom_sf(data = area, fill = "grey95", colour = "grey35") +
  geom_sf(data = cases, colour = "#b2182b", alpha = 0.55, size = 0.8) +
  coord_sf(datum = NA) +
  labs(title = "Simulated case locations, January–June 2024",
       x = "Easting (m)", y = "Northing (m)") +
  theme_minimal()
```
Answer:

1) The file contains 500 simulated dengue cases. Each row represents one reported case, including its location, neighbourhood, age group, onset date, and report date.

2) I first assigned the case locations their latitude and longitude coordinate system (EPSG:4326), then projected both the cases and the boundary into UTM Zone 18N (EPSG:32618). The analysis cannot be done in degrees because degrees are angles, not lengths: a degree of longitude covers a different ground distance than a degree of latitude, so distances, areas and intensities in degrees would not be in consistent units.

3) The polygon window contains 500 events and covers approximately 190.35 square kilometres. Its overall intensity is approximately 2.63 cases per square kilometre over the six-month period.

4) The cases’ bounding rectangle covers approximately 212.95 square kilometres and gives about 2.35 cases per square kilometre, roughly 10.6% lower than the polygon result. The case count is still 500; the intensity changes because the rectangle includes a larger area. The window defines where cases could have been observed and affects the comparisons used later, so I use the study-area polygon for the remaining analyses.

5) The map shows the case locations within the study-area boundary, with visible concentrations in several parts of the area.

---

## Part B --- Quadrat analysis (3 points)

1. Lay a 6 × 6 grid over the window, map the counts, and compute the mean,
   variance and **variance-to-mean ratio** from the counts by hand --- not from a
   canned function. Show the arithmetic.
2. State what the VMR is being compared against, and why: under complete spatial
   randomness the counts are Poisson, and a Poisson distribution has variance
   equal to its mean. Report where your value sits relative to 1.
3. Repeat the VMR for at least four grid resolutions and plot VMR against cells
   per side. Three or four sentences: what does the shape of that curve say about
   the claim "this pattern is clustered"?

### Starting code


```{r}
#| label: quadrats
#| fig-width: 7
#| fig-height: 6
qc <- quadratcount(pp, nx = 6, ny = 6)
cnt <- as.vector(qc)
k <- length(cnt)
q_mean <- sum(cnt) / k
q_variance <- sum((cnt - q_mean)^2) / (k - 1)
q_vmr <- q_variance / q_mean
stopifnot(sum(cnt) == npoints(pp))
print(c(k = k, total = sum(cnt), mean = q_mean,
        sample_variance = q_variance, VMR = q_vmr))
plot(qc, main = "Case counts in a 6 × 6 grid")
```

```{r}
#| label: quadrat-sensitivity
#| fig-width: 7
#| fig-height: 4
grid_n <- c(3, 4, 6, 8, 12)
vmr <- vapply(grid_n, function(g) {
  v <- as.vector(quadratcount(pp, nx = g, ny = g))
  mu <- sum(v) / length(v)
  (sum((v - mu)^2) / (length(v) - 1)) / mu
}, numeric(1))
vmr_table <- data.frame(cells_per_side = grid_n, VMR = vmr)
knitr::kable(vmr_table, digits = 2)
ggplot(vmr_table, aes(cells_per_side, VMR)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(colour = "#2166ac") +
  geom_point(size = 2, colour = "#2166ac") +
  scale_x_continuous(breaks = grid_n) +
  labs(x = "Cells per side", y = "Variance / mean",
       title = "Quadrat results change with grid resolution") +
  theme_minimal()
```

Answer:

1) The 6 × 6 grid has 36 cells containing 500 cases. The mean is 500 divided by 36, or approximately 13.889 cases per cell. The sum of squared differences from that mean is approximately 20,359.56; dividing by 35 gives a sample variance of about 581.702. The variance divided by the mean gives a VMR of about 41.883.

2) Under homogeneous complete spatial randomness, counts in equal-area cells have a Poisson distribution, whose variance equals its mean. The comparison value for the VMR is therefore 1. The observed value of about 41.88 is far above 1, indicating strongly uneven counts across the grid. Cells cut by the polygon have unequal usable areas, so this is a descriptive comparison.

3) The VMR is approximately 33.00, 35.17, 41.88, 32.68, and 30.23 for grids with 3, 4, 6, 8, and 12 cells per side, respectively. It peaks at the 6 × 6 grid and remains far above 1 at every resolution. This supports the clustering description across several grid sizes, although the apparent strength changes with cell size. Grid placement and cells cut by the boundary also affect the result.


---

## Part C --- Nearest-neighbour statistics (3 points)

1. Compute `nndist()`, report the observed mean, and compute the CSR expectation
   $E(d) = 1/(2\sqrt{\lambda})$ **by hand** from your intensity. Report the
   Clark--Evans ratio $R$ and state the direction of the departure.
2. Run `clarkevans.test()` with a Monte Carlo null. The window is an irregular
   polygon, so the Donnelly correction is unavailable --- use
   `correction = "cdf"` with `method = "MonteCarlo"`. One sentence on why an edge
   correction is needed at all.
3. Plot the G function with a simulation envelope. Two or three sentences on what
   G shows that the single ratio $R$ does not.

### Starting code

```{r}
#| label: nearest-neighbour
nnd <- nndist(pp)
lambda <- npoints(pp) / area.owin(Window(pp))  # cases / m^2
d_expected <- 1 / (2 * sqrt(lambda))
R_ce <- mean(nnd) / d_expected
print(c(observed_mean_m = mean(nnd), CSR_expected_m = d_expected,
        Clark_Evans_R = R_ce))
ce_test <- clarkevans.test(pp, correction = "cdf",
                          alternative = "clustered",
                          method = "MonteCarlo", nsim = 99)
ce_test
```
```{r}
#| label: g-envelope
#| fig-width: 7
#| fig-height: 5
g_env <- envelope(pp, Gest, nsim = 99, verbose = FALSE)
plot(g_env, main = "Nearest-neighbour distribution G(r)", legend = FALSE)
```
Answer:

1) The observed mean nearest-neighbour distance is approximately 215.5 m. The polygon intensity is about 0.000002627 cases per square metre; using this in the stated CSR formula gives an expected distance of about 308.5 m. The Clark–Evans ratio is approximately 0.699. Because it is below 1, the cases have closer nearest neighbours than expected under CSR, in the direction of clustering.

2) I tested the clustering direction using 99 CSR simulations and the CDF edge correction. With the CDF edge correction the ratio is R = 0.655, slightly further below 1 than the uncorrected 0.699. The Monte Carlo p-value is 0.01, the smallest value resolvable with 99 simulations, so p ≤ 0.01: none of the simulated CSR patterns was as clustered as the observed cases. Edge correction is needed because a case near the boundary might have a closer neighbour outside the area that cannot be observed in these data.

3) The observed G curve lies above the upper pointwise CSR simulation band for approximately 20–430 m. Around 100 m, nearly half the cases have a nearest neighbour within that distance, much more than expected under CSR. The G curve shows which distances drive the departure, while the Clark–Evans ratio reduces the pattern to one average.


---

## Part D --- Ripley's K across scales (4 points)

1. Estimate $K(r)$ with an edge correction, then plot the centred L function
   $\hat{L}(r) - r$ against $r$.
2. Add a **simultaneous** envelope (`global = TRUE`) from at least 99 CSR
   simulations. State the distances over which the curve leaves the band.
3. Read a **characteristic scale** off the curve --- the $r$ at which the
   departure peaks --- and translate it into something a public health officer
   would recognise: a few city blocks, a neighbourhood, half the city.
4. Three or four sentences: $K$ assumes the intensity is constant across the
   window. Looking at your map from Part A, is that plausible here? Say what the
   consequence is for reading the curve --- what else, besides interaction
   between cases, could put it above the envelope.

::: {.callout-warning}
`global = TRUE` with `nsim = 99` takes a minute or two. Develop with
`nsim = 19`, raise it once, and do not re-render in a loop.
:::

### Starting code

```{r}
#| label: l-estimate-and-envelope
#| fig-width: 7
#| fig-height: 5
r_grid <- seq(0, 3000, by = 25)
K_hat <- Kest(pp, r = r_grid, correction = "isotropic")
L_hat <- Lest(pp, r = r_grid, correction = "isotropic")
plot(L_hat, . - r ~ r, main = "Centred L(r), isotropic correction",
     legend = FALSE)
abline(h = 0, col = "grey50", lty = 2)

# The 99 CSR simulations give a global/simultaneous envelope over r_grid.
l_env <- envelope(pp, Lest, r = r_grid, correction = "isotropic",
                  nsim = 99, global = TRUE, verbose = FALSE)
plot(l_env, . - r ~ r, main = "Centred L with global CSR envelope",
     legend = FALSE)
abline(h = 0, col = "grey50", lty = 2)

above_r <- l_env$r[is.finite(l_env$obs) & is.finite(l_env$hi) &
                     l_env$obs > l_env$hi]
below_r <- l_env$r[is.finite(l_env$obs) & is.finite(l_env$lo) &
                     l_env$obs < l_env$lo]
peak_i <- which.max(replace(l_env$obs - l_env$r,
                           !is.finite(l_env$obs - l_env$r), -Inf))
peak_r <- l_env$r[peak_i]
print(list(
  above_envelope_m = if (length(above_r)) range(above_r) else NULL,
  below_envelope_m = if (length(below_r)) range(below_r) else NULL,
  peak_distance_m = peak_r,
  peak_L_minus_r_m = l_env$obs[peak_i] - peak_r
))

```
Answer:

1) I estimated K(r) using the isotropic edge correction and plotted centred L against distance. The curve is positive across nearly the entire tested range, indicating more spatial concentration than expected under homogeneous complete spatial randomness (CSR).

2) I constructed a simultaneous CSR envelope using 99 simulations. The observed curve is above the upper band from 25 m through 3,000 m, the largest distance tested. It does not fall below the lower band at any tested distance.

3) The centred L curve peaks at a distance of 775 m. At that distance, the plotted value of L(r) minus r is 1,566.25 m. A 775 m radius represents roughly a neighbourhood-scale pattern; 1,566.25 m is the height of the departure on the graph, not the cluster radius.

4) A constant case intensity across the polygon is not plausible from the Part A map, which shows dense concentrations in some locations and sparse areas elsewhere. Homogeneous CSR assumes the same background intensity everywhere, so variation in that background can also raise the L curve above its envelope. Differences in population density, local conditions, or reporting could produce such variation without direct interaction between cases. This result shows a strong departure from homogeneous CSR, but does not establish what caused the pattern.

---

## Part E --- Kernel density estimation and bandwidth choice (4 points)

K tells you *whether* and *at what distance*; it does not tell you *where*. That
is what a kernel intensity surface is for.

1. Estimate the intensity with `density.ppp()` and convert to **cases per km²**.
   Map it with the cases drawn on top.
2. Produce the same surface at **three bandwidths** --- one clearly too small,
   one clearly too large, and one you would defend --- on a **common colour
   scale** so the panels are comparable. Two or three sentences on what changes
   and what does not.
3. Compare at least two automatic selectors (`bw.diggle()`, `bw.ppl()`,
   `bw.scott()`). Report the values in metres, say which you adopt and why, and
   note whether your answer to step 2 would change if you adopted a different one.
4. Three or four sentences on what this map is and is not. Where is the intensity
   highest? What would you need in order to turn this surface into a statement
   about **risk** rather than about counts, and what is the name of the quantity
   you would compute instead?

::: {.callout-tip}
## What earns full marks here
Step 4, and the common colour scale in step 2. Two surfaces at different
bandwidths with independently stretched legends look like different data; a
reader cannot compare them, and neither can you.
:::

### Starting code

```{r}
#| label: bandwidths
bandwidths <- c(Diggle = as.numeric(bw.diggle(pp)),
                PPL = as.numeric(bw.ppl(pp)),
                Scott = as.numeric(bw.scott(pp, isotropic = TRUE)))
knitr::kable(data.frame(selector = names(bandwidths),
                        sigma_m = unname(bandwidths)), digits = 1)
h <- unname(bandwidths["Diggle"])
stopifnot(is.finite(h), h > 0)
```

```{r}
#| label: main-kde
#| fig-width: 7
#| fig-height: 6
kde <- density(pp, sigma = h, eps = 50) * 1e6
plot(kde, main = paste0("Case intensity, σ = ", round(h), " m"),
     ribargs = list(las = 1))
plot(Window(pp), add = TRUE, border = "grey30")
points(pp, pch = 16, cex = 0.25, col = rgb(0, 0, 0, 0.4))
mtext("Estimated cases per km² over the six-month record", side = 1, line = 3)
```

```{r}
#| label: compare-bandwidths
#| fig-width: 12
#| fig-height: 4.5
sigmas <- c(small = min(150, h / 4), chosen = h, large = max(3500, h * 4))
surf <- lapply(sigmas, function(sigma) density(pp, sigma = sigma, eps = 50) * 1e6)
zmax_e <- max(vapply(surf, function(im) max(im$v, na.rm = TRUE), numeric(1)))
oldpar <- par(mfrow = c(1, 3), mar = c(2, 2, 3, 3))
for (i in seq_along(surf)) {
  plot(surf[[i]], main = paste0(names(sigmas)[i], ": σ = ",
                                round(sigmas[i]), " m"), zlim = c(0, zmax_e))
  plot(Window(pp), add = TRUE, border = "grey35")
}
par(oldpar)
```
Answer:

1) The map estimates case intensity over the six-month period in cases per square kilometre, with the individual case locations drawn on top.

2) At 42 metres, the surface breaks into very small, sharp peaks around individual groups of cases. At the chosen 169 metres, the main concentrations remain visible without as much fine detail. At 3500 metres, the surface is so smooth that the local hotspots nearly disappear. The underlying case locations do not change; only how broadly their intensity is spread changes. All three panels use the same colour scale, so their colours can be compared directly.

3) The Diggle selector gives 168.7 metres, the PPL selector gives 469.9 metres, and the Scott selector gives 1250.5 metres. I adopt the Diggle bandwidth because it shows the distinct local concentrations without reducing the map to isolated points. PPL would make those concentrations broader, and Scott would smooth them more strongly, but neither would change the basic conclusion that cases are concentrated in particular areas.

4) This is a map of case intensity, meaning estimated case counts per unit of area; it is not a map of each resident’s chance of getting dengue. Intensity is highest around Floralia, Siloé, and Aguablanca in the pooled six-month map. A place with more residents could have more cases even if individual risk were no higher. To discuss risk, I would need the population at risk (or a set of controls) for the same places and period. Dividing the case intensity by the population intensity gives a relative risk surface, or equivalently a smoothed incidence rate such as cases per 1,000 residents.

---

## Part F --- Space-time kernel density (3 points)

A single intensity surface pools six months. An outbreak is a hotspot that
exists in some weeks and not others, and pooling hides exactly that.

1. Split the cases into **monthly** patterns and estimate a kernel surface for
   each, using **one bandwidth for all months** and **one colour scale for all
   months**. Display them as small multiples.
2. Two or three sentences on why both of those constraints are necessary, and
   what a reader would wrongly conclude if each panel were scaled to its own
   maximum.
3. Name the months and locations where the intensity is high, and say which
   hotspots persist across months and which appear and disappear.
4. Build a **space-time cube**: an interactive 3D scatter with easting, northing
   and day-of-study on the three axes, coloured by neighbourhood. Rotate it until
   the structure is visible, then set a viewing angle that makes your point and
   keep it. One or two sentences on what the cube shows that the small multiples
   do not, and vice versa --- look in particular for the difference between a
   compact blob and a vertical column.

::: {.callout-warning}
## plotly renders in HTML only
The cube is an HTML widget: it appears in an HTML render and is silently dropped
from a PDF one. The two chunks below are gated on `knitr::is_html_output()`, so
whichever format you render gets a working figure. If you publish the HTML, keep
`embed-resources: true` --- otherwise the widget lives in a `_files/` folder that
has to travel with the page.
:::

::: {.callout-note}
## If you want the proper estimator
Steps 1--3 are a discretised space-time KDE: one spatial bandwidth, and a
temporal "bandwidth" equal to one month, applied as a hard cut. The continuous
version smooths in time as well, with a separate temporal bandwidth, and lives
in the `sparr` package --- `spattemp.density()` with `h` and `lambda`, then
`spattemp.slice()` to pull out time slices. Using it is worth up to **2 bonus
points** if the monthly version is already complete and correct. Do not start
here.
:::

### Starting code

```{r}
#| label: f-monthly-maps
#| fig-width: 12
#| fig-height: 8

cases$month <- format(cases$onset_date, "%Y-%m")
months <- sort(unique(cases$month))

# Use the Diggle bandwidth chosen in Part E for every month.

by_month <- lapply(months, function(m) {
  sub <- cases[cases$month == m, ]
  coords <- st_coordinates(sub)
  monthly_pp <- ppp(coords[, 1], coords[, 2],
                    window = Window(pp))
  density(monthly_pp, sigma = h, eps = 50) * 1e6
})

# Find one maximum for the colour scale of all six maps.
zmax_f <- max(vapply(by_month,
                    function(im) max(im$v, na.rm = TRUE),
                    numeric(1)))

oldpar <- par(mfrow = c(2, 3), mar = c(2, 2, 3, 3))
for (i in seq_along(by_month)) {
  plot(by_month[[i]],
       main = paste0(months[i], " (n = ",
                     sum(cases$month == months[i]), ")"),
       zlim = c(0, zmax_f))
  plot(Window(pp), add = TRUE, border = "grey35")
}
par(oldpar)

xy <- st_coordinates(cases)
cube <- data.frame(
  E = xy[, 1],
  N = xy[, 2],
  day = as.numeric(cases$onset_date - min(cases$onset_date)),
  hood = cases$neighborhood,
  onset = cases$onset_date
)
```

```{r}
#| label: f-cube-html
#| eval: !expr knitr::is_html_output()

plotly::plot_ly(
  cube,
  x = ~E, y = ~N, z = ~day,
  color = ~hood,
  colors = hcl.colors(length(unique(cube$hood)), "Dark 3"),
  text = ~paste(hood, onset),
  hoverinfo = "text",
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, opacity = 0.7)
) |>
  plotly::layout(scene = list(
    xaxis = list(title = "Easting (m)"),
    yaxis = list(title = "Northing (m)"),
    zaxis = list(title = "Day of study"),
    camera = list(eye = list(x = 1.7, y = -1.6, z = 0.9))
  ))
```

```{r}
#| label: f-cube-static
#| eval: !expr isFALSE(knitr::is_html_output())

scatterplot3d::scatterplot3d(cube$E, cube$N, cube$day, pch = 16, cex.symbols = 0.4,
                             color = as.integer(factor(cube$hood)),
                             xlab = "easting (m)", ylab = "northing (m)", zlab = "day")
```
Answer:


1) I split the 500 cases by onset month (January–June 2024) and estimated one kernel surface per month, using the Diggle bandwidth from Part E (σ ≈ 169 m) for every month and a single colour scale from zero to the highest monthly value. Each panel shows estimated cases per km² for that month, with the monthly case count in its title.

2) The bandwidth controls how peaked a surface looks, so a smaller σ in one month would make its hotspots look sharper and higher even with identical cases; differences between panels would then reflect smoothing rather than data. A shared colour scale makes each colour mean the same intensity in every month. If each panel were scaled to its own maximum, every month would show a dark hotspot, a quiet month would look like an outbreak, and the timing of the outbreaks would disappear.

3) Intensity is highest in Floralia in February (72 cases), Aguablanca in March (105) and Siloé in May (84). Each is a transient hotspot confined to a single month: in every other month these neighbourhoods record only 1–8 cases. El Retiro is the persistent concentration: it has 13–24 cases in every month, so it appears at a moderate intensity in all six panels rather than spiking in any one.

4) I set the camera so that time runs up the vertical axis. From that angle the three outbreak neighbourhoods appear as compact blobs at different heights (Floralia lowest, then Aguablanca, then Siloé), while El Retiro forms a vertical column through the whole study period. The cube shows exact onset timing, so an outbreak that straddles a month boundary stays one blob rather than being split across two panels. The small multiples show intensity on a common scale and the spatial extent of each hotspot, which is hard to judge from overlapping points in 3D.

---

## Part G --- The space-time K function (4 points)

Spatial clustering and space-time clustering are different claims. A
neighbourhood with chronically high incidence is clustered in space in every
month; an outbreak is clustered in space *and* time together. Diggle's space-time
K function separates them by comparing the joint function with the product of the
two marginals:

$$
D(s, t) = K_{st}(s, t) - K_s(s)\,K_t(t)
$$

$D = 0$ means space and time are independent --- cases may cluster spatially,
temporally, or both, but not *jointly*. $D > 0$ is space-time interaction.

1. Build the inputs `splancs` expects: a two-column matrix of projected
   coordinates, a numeric time vector (days since the first onset), and the
   boundary as a matrix of vertices.
2. Compute $K_s$, $K_t$ and $K_{st}$ with `stkhat()`. Justify your choice of
   spatial distances `s` and time lags `tm` in one sentence each --- what scales
   are plausible for a mosquito-borne disease?
3. Plot **both** $D(s,t)$ and the standardised version
   $D_0(s,t) = K_{st}/(K_s K_t) - 1$. One or two sentences on why the two
   surfaces have different shapes: $K$ functions are cumulative, so $D$
   accumulates and then flattens, while $D_0$ measures the *proportional* excess
   and falls away once you leave the scale of the clustering. This is the same
   move as plotting $\hat{L}(r) - r$ instead of $\hat{K}(r)$ in Part D.
4. A surface is hard to read against nothing. Build a **permutation envelope**:
   shuffle the onset dates across the fixed locations, recompute $D_0$, repeat 99
   times, and take the upper 97.5% quantile at each $(s, t)$. Then plot $D_0$
   against $s$ as a curve per time lag, with its envelope --- the space-time
   analogue of the envelope you used for $L$. Shuffling the dates destroys any
   space-time interaction while leaving both marginal patterns exactly as
   observed, which is the correct null here.
5. Read the peak off those curves --- a distance and a lag --- and say what the
   pair means for this disease.
6. Run `stmctest()` for a single global p-value. It sums the residuals
   $\hat{D}(s,t)$ over the whole grid into one statistic $U$, recomputes it under
   permuted dates, and returns the observed value as `$t0` and the permuted ones
   as `$t`. Report the p-value as
   $(1 + \#\{t \geq t_0\}) / (1 + B)$ and plot the null distribution with the
   observed value marked. One sentence on what this test adds to the curves in
   step 4, and what it cannot tell you.
7. Five or six sentences. Tabulate cases by neighbourhood and month. At least one
   neighbourhood contributes to the spatial clustering of Part D without
   contributing to the space-time interaction here. Identify it, explain what
   distinguishes it from the others, and say what each of the two patterns would
   mean for someone deciding where to send a vector-control team this week.

::: {.callout-warning}
## Never report $p = 0$
`mean(mc$t >= mc$t0)` returns 0 when no permutation beats the observed value,
and 0 is not a p-value --- with $B$ permutations you cannot resolve anything
below $1/(B+1)$. Always add the observed value to both counts, as above. With
999 permutations the smallest reportable value is 0.001, and the honest phrasing
is "$p \leq 0.001$", not "$p < 0.0001$".
:::

::: {.callout-note}
`splancs` is old and its arguments are terse: `stkhat(pts, times, poly, tlimits,
s, tm)`, where `tlimits` is the time range, `s` the spatial distances and `tm`
the time lags. `?stkhat` and `?stmctest` are short. A **Knox test** on a distance
and time threshold of your choosing is an acceptable substitute if you defend
both thresholds --- say explicitly if you go that route.
:::

### Starting code

```{r}
#| label: g-space-time-analysis
#| fig-width: 11
#| fig-height: 5

# 1. Prepare projected locations, onset times, and boundary vertices
xy <- st_coordinates(cases)
pts <- as.points(xy[, 1], xy[, 2])
times <- as.numeric(cases$onset_date - min(cases$onset_date))

bdry <- st_coordinates(area)[, 1:2, drop = FALSE]
if (all(bdry[1, ] == bdry[nrow(bdry), ])) {
  bdry <- bdry[-nrow(bdry), , drop = FALSE]
}

# 2. Estimate spatial, temporal, and joint space-time functions
s <- c(seq(50, 1000, by = 50), seq(1200, 3000, by = 200))
tm <- seq(3, 69, by = 3)
tlimits <- c(0, max(times))

stk <- stkhat(pts, times, bdry, tlimits, s, tm)

# 3. Calculate and plot raw and proportional excess
denom <- outer(stk$ks, stk$kt)
D <- stk$kst - denom
D0 <- stk$kst / denom - 1

oldpar <- par(mfrow = c(1, 2))
persp(s, tm, D, theta = 30, phi = 25,
      xlab = "Distance (m)", ylab = "Lag (days)",
      zlab = "Raw excess", col = "#91bfdb")
persp(s, tm, D0, theta = 30, phi = 25,
      xlab = "Distance (m)", ylab = "Lag (days)",
      zlab = "Proportional excess", col = "#fc8d59")
par(oldpar)

# 4. Shuffle onset times 99 times to make a pointwise envelope
B <- 99
perm <- array(NA_real_, dim = c(length(s), length(tm), B))

for (b in seq_len(B)) {
  shuffled <- stkhat(pts, sample(times), bdry, tlimits, s, tm)
  perm[, , b] <- shuffled$kst /
    outer(shuffled$ks, shuffled$kt) - 1
}

hi <- apply(perm, c(1, 2), quantile,
            probs = 0.975, na.rm = TRUE)

lag_index <- match(c(3, 9, 21, 42), tm)

matplot(s, D0[, lag_index, drop = FALSE],
        type = "l", lty = 1, lwd = 2, col = 1:4,
        xlab = "Distance (m)",
        ylab = "Proportional excess",
        ylim = range(c(D0[, lag_index],
                       hi[, lag_index]), finite = TRUE))
matlines(s, hi[, lag_index, drop = FALSE],
         lty = 3, col = 1:4)
abline(h = 0, col = "grey60")
legend("topright",
       legend = paste(tm[lag_index], "day lag"),
       col = 1:4, lty = 1, lwd = 2, bty = "n")

# 5. Find the largest gap above the envelope on the plotted curves
excess <- D0[, lag_index, drop = FALSE] -
  hi[, lag_index, drop = FALSE]
peak <- which(excess == max(excess, na.rm = TRUE),
              arr.ind = TRUE)[1, ]

peak_distance <- s[peak[1]]
peak_lag <- tm[lag_index[peak[2]]]

print(c(peak_distance_metres = peak_distance,
        peak_lag_days = peak_lag))

# 6. Global permutation test using 999 shuffled datasets
mc <- stmctest(pts, times, bdry, tlimits, s, tm,
               nsim = 999, quiet = TRUE)

p_global <- (1 + sum(mc$t >= mc$t0)) /
  (1 + length(mc$t))

print(c(observed_statistic = mc$t0,
        permutations = length(mc$t),
        global_p_value = p_global))

hist(mc$t, breaks = 30,
     xlim = range(c(mc$t, mc$t0)),
     xlab = "Statistic after shuffling dates",
     main = "Global permutation test")
abline(v = mc$t0, col = "firebrick", lwd = 2)

# 7. Cases by neighbourhood and month
hood_month <- table(cases$neighborhood, cases$month)
knitr::kable(as.data.frame.matrix(hood_month),
             caption = "Cases by neighbourhood and month")
```
Answer:


1) I used the cases’ projected easting and northing coordinates, measured onset times as days since the first onset, and supplied the study-area boundary as polygon vertices.

2) The spatial distances run from 50 metres to 3 kilometres, with finer steps near the origin, because the *Aedes aegypti* mosquito rarely flies more than a few hundred metres, so this range brackets everything from a single household to several neighbourhoods. The time lags run from 3 to 69 days because the extrinsic incubation in the mosquito (about 8–12 days) plus the intrinsic incubation in people (about 4–7 days) puts roughly two to three weeks between linked cases, so this range covers several transmission generations within the six-month record.

3) The raw excess surface accumulates as larger distances and time lags include more pairs of cases. The proportional excess surface highlights where the joint concentration is strongest relative to the separate spatial and temporal patterns, then generally declines at broader scales. A positive surface alone does not establish statistical significance.

4) The solid curves are the observed proportional excess, and the dotted curves are the upper pointwise limits from 99 shuffled-date datasets. The observed curves lie above their corresponding limits across the plotted distances, with the strongest separation at short time lags. Shuffling dates preserves where cases occurred and the overall timing of cases while breaking the link between particular locations and onset dates.

5) Among the four displayed time lags, the largest gap above the upper limit occurs at 400 metres and 3 days. This suggests particularly strong space-time association among cases separated by a few city blocks and only a few days. A 400 m distance is consistent with the flight range of the vector plus local human movement. A 3-day lag is shorter than one transmission cycle, so it more likely reflects cases infected at around the same time by the same local mosquito population than one case infecting the next. Because 3 days is the shortest lag I examined, the strongest association may lie at even shorter lags. The pattern does not by itself establish how transmission occurred.

6) The observed global statistic is approximately 723.4 billion, far to the right of all 999 shuffled results in the histogram. None of those results reached the observed value, so the corrected global p-value is (1 + 0) / (1 + 999) = 0.001, which I report as p ≤ 0.001. This provides strong evidence of an association between case location and onset time, but the global test alone cannot identify the responsible neighbourhood or its characteristic distance and lag.

7) El Retiro recorded 20, 13, 24, 14, 13, and 13 cases from January through June, so its cases are relatively persistent across the study. It can contribute to spatial clustering without producing a sharp surge in one particular month. In contrast, Floralia recorded 72 cases in February, Aguablanca 105 in March, and Siloé 84 in May. Those concentrated monthly surges help explain the joint space-time pattern. Persistent cases in El Retiro could warrant ongoing surveillance and prevention, while a new local surge would call for a more immediate, targeted vector-control response. These simulated historical records cannot by themselves determine where a real team should go this week.

---

## Part H --- Rendering and publication (1 point)

The `.qmd` is the submission, but the rendered page is what anyone else can
actually read --- and by the end of the semester this repository is your
portfolio.

1. Render to **HTML** from a clean session, with the seed set and every package
   loaded explicitly, then commit both the `.qmd` and the rendered output to your
   semester repository, in the folder your site is built from.
2. Push, wait for GitHub Pages to rebuild, and **open the published URL yourself**
   to check it. A page that renders locally and 404s online is not submitted.
3. Put the URL of the published page in your document header and in your Course
   Site submission.

The point is for a page that works: rendered, pushed, and reachable at the URL
you supply.

::: {.callout-warning}
## Two things that break the published page
`embed-resources: true` is already set in the header --- keep it. Without it the
`plotly` cube in Part F lives in a `_files/` folder that must be committed
alongside the HTML, and forgetting it gives a page with a blank space where the
cube should be.

Relative paths to `data/` work locally because your working directory is the
`.qmd`'s folder. They will also work on the site *as long as you commit the data
files*. Check the live page, not the local preview.
:::

Answer:

I rendered this document to HTML from a clean R session, with the seed set in the setup chunk and every package loaded explicitly, and kept `embed-resources: true` so the plotly cube is embedded in the page. I committed the `.qmd`, the rendered HTML and the `data/` folder to my semester repository, pushed, and checked the live GitHub Pages URL given in the header after the site rebuilt.

---

## AI use {.unnumbered}

End your document with a short **AI use note**: which tools, what for, at which
stage. Two or three sentences.

AI use is unrestricted in this course and costs you nothing, but you must be able
to explain every line you hand in. The note carries no points --- it is simply
expected, the way a methods section is.

**AI use note.** I used Claude (Anthropic) after writing the code and first-draft answers, to cross-check the arithmetic and interpretation in each part and to tighten the wording of several written answers, especially in Parts F and G. 

---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct import, CRS and window; `ppp` built and explained; effect of the window argued | 3 |
| **B** | VMR computed from counts; benchmark of 1 explained; resolution sweep interpreted | 3 |
| **C** | $E(d)$ derived and $R$ interpreted; Monte Carlo test correct; G compared with $R$ | 3 |
| **D** | Centred L with a simultaneous envelope; characteristic scale read and translated; stationarity assumption discussed | 4 |
| **E** | Surface in interpretable units; three bandwidths on a common scale; selectors compared; density-is-not-risk argument | 4 |
| **F** | Monthly surfaces with common bandwidth and scale; persistent vs transient hotspots named; space-time cube | 3 |
| **G** | Space-time K computed; $D_0$ with a permutation envelope; peak distance and lag read off; the spatial-only neighbourhood identified and the operational difference explained | 4 |
| **H** | Clean render to HTML, published to GitHub Pages; working URL supplied | 1 |
| | **Total** | **25** |

Writing counts for roughly 15% within each part --- more than in Exercise 1,
because more of this lab is interpretation. A submission that is all code and no
sentences cannot score above about 50%.

---

## Getting help

- Office hours, or email with the code and the exact error message.
- `splancs` in Part G is the one place where the documentation is thin. Start
  early enough that you can ask.
- Stuck for more than thirty minutes on something mechanical: ask rather than
  persevere.
